Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-09-27 analysis/paper/paper.Rmd
Removed old project file name, added river classification to data files and updated code

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
vis.dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:27),c(78:428))

spec.vd <- 
      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:27)

vis.light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:27),c(78:428))

spec.vl <- 
  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:27)

vis.translucent <- Points.nir %>% 
  filter(hue == "Translucent") %>% 
  select(c(1:27),c(78:428))

spec.vt <- 
  as_spectra(vis.translucent, name_idx = 1, meta_idxs = 2:27)

vis.white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:27),c(78:428))

spec.vw <- 
  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:27)

@@

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:27),c(678:2178))

#create a spectra object for the dark material, and then repeat for light, white and translucent groups
spec.d <- 
  as_spectra(dark, name_idx = 1, meta_idxs = 2:27)

light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:27),c(678:2178))

spec.l <- 
  as_spectra(light, name_idx = 1, meta_idxs = 2:27)

translucent <- Points.nir %>% 
  filter(hue == "Translucent") %>% 
  select(c(1:27),c(678:2178))

spec.t <- 
  as_spectra(translucent, name_idx = 1, meta_idxs = 2:27)

white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:27),c(678:2178))

spec.w <- 
  as_spectra(white, name_idx = 1, meta_idxs = 2:27)

#Prepare a layout with 1 column and four rows
par(mfrow=c(4,1))
#Plot the mean spectrum for each spectra group above
plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark")

plot(mean(spec.l), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")

plot(mean(spec.t), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Translucent")

plot(mean(spec.w), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White")

@@

#perform PCA with SNV normalization and mean-center
nir.pca <-
  prcomp(Points.nir[,c(678:2178)], center = TRUE, scale = FALSE)

@@

@@

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on different types of shale, as well as samples of diabase, apatite, orthoclase, and blue quartz. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group. Considering the diagenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.

@@

```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue. The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the dark group"}
#Read and create new file for chemospec
#raman <-matrix2SpectraObject(
  #gr.crit = c("Translucent", "White", "Dark", "Light" ),
  #gr.cols = c("purple", "blue", "black", "yellow"),
  #freq.unit ="Wavenumber (cm-1)",
  #int.unit ="Intensity",
  #descrip ="Bifacial points measurements",
  #in.file = "./analysis/data/derived_data/raman_baseline_corrected_transposed_20210908.csv",
  #sep = ";",
  #dec = ",",
 # chk = TRUE,
 # out.file = "raman")

#surveySpectra(raman, by.gr = TRUE , method = "sem", main = "mean")

@@

spec.rd <- 
  as_spectra(r.dark, name_idx = 1, meta_idxs = 2:28)

@@

spec.rl <- 
  as_spectra(r.light, name_idx = 1, meta_idxs = 2:28)

@@

spec.rt <- 
  as_spectra(r.translucent, name_idx = 1, meta_idxs = 2:28)

@@

spec.rw <- 
  as_spectra(r.white, name_idx = 1, meta_idxs = 2:28)

@@

#perform PCA with SNV normalization and mean-center
raman.pca <-
  prcomp(standardNormalVariate(X = Points.raman[,c(29:320)]), center = TRUE, scale = FALSE)

@@

>>>>

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
vis.dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:28),c(79:429))

spec.vd <- 
      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)

vis.light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:28),c(79:429))

spec.vl <- 
  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)

vis.translucent <- Points.nir %>% 
  filter(hue == "Translucent") %>% 
  select(c(1:28),c(79:429))

spec.vt <- 
  as_spectra(vis.translucent, name_idx = 1, meta_idxs = 2:28)

vis.white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:28),c(79:429))

spec.vw <- 
  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)

@@

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:28),c(679:2179))

#create a spectra object for the dark material, and then repeat for light, white and translucent groups
spec.d <- 
  as_spectra(dark, name_idx = 1, meta_idxs = 2:28)

light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:28),c(679:2179))

spec.l <- 
  as_spectra(light, name_idx = 1, meta_idxs = 2:28)

translucent <- Points.nir %>% 
  filter(hue == "Translucent") %>% 
  select(c(1:28),c(679:2179))

spec.t <- 
  as_spectra(translucent, name_idx = 1, meta_idxs = 2:28)

white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:28),c(679:2179))

spec.w <- 
  as_spectra(white, name_idx = 1, meta_idxs = 2:28)

#Prepare a layout with 1 column and four rows
par(mfrow=c(4,1))
#Plot the mean spectrum for each spectra group above
plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
     xlab = "Wavelength (nm)", main = "Average - Dark")

plot(mean(spec.l), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
     xlab = "Wavelength (nm)", main = "Average - Light")

plot(mean(spec.t), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
     xlab = "Wavelength (nm)", main = "Average - Translucent")

plot(mean(spec.w), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
     xlab = "Wavelength (nm)", main = "Average - White")

@@

#perform PCA with SNV normalization and mean-center
nir.pca <-
  prcomp(Points.nir[,c(679:2179)], center = TRUE, scale = FALSE)

@@

#NIR LOADING LINES
#par(mfrow=c(3,1))
#sapply(1:3, function(comp){
#  plot(1000:2500, nir.pca$rotation[,comp], type='l', lwd=2,
#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(1000,2500,6))})

@@

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group. Considering the diagenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.

@@

```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue. The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the dark group"}

@@

spec.rd <- 
  as_spectra(r.dark, name_idx = 1, meta_idxs = 2:29)

@@

spec.rl <- 
  as_spectra(r.light, name_idx = 1, meta_idxs = 2:29)

@@

spec.rt <- 
  as_spectra(r.translucent, name_idx = 1, meta_idxs = 2:29)

@@

spec.rw <- 
  as_spectra(r.white, name_idx = 1, meta_idxs = 2:29)

@@

#perform PCA with SNV normalization and mean-center
raman.pca <-
  prcomp(standardNormalVariate(X = Points.raman[,c(30:321)]), center = TRUE, scale = FALSE)

@@

#RAMAN LOADING LINES
#par(mfrow=c(2,2))
#sapply(1:4, function(comp){
#  plot(row.names(raman.pca$rotation), raman.pca$rotation[,comp], type='l', lwd=2, 
#       main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(92.88,1798.3,6))})
